Sentence Generation as Planning with Probabilistic LTAG\\

### 2010-05-21 trunk/Papers/pcrisp-10/pcrisp.tex
einmal ganz drueber

We now extend {\sc crisp} to statistical generation ({\sc pcrisp}). The basic idea is to add a statistical grammar model while leaving the sentence generation mechanism untouched. This way we can select the highest scoring derivation, that satisfies all constraints (grammaticality, expresses the communicative goal, use unambiguous referring expressions etc.).

@@

While we leave the basic sentence generation mechanism intact, we need to modify the concrete formulation of {\sc crisp} planning operators, as each of them now corresponds to some probability event. Likewise, we need to take the step from classical planning to metric planning systems, that can use the probabilities.
In metric planning \cite{fox2002}, instead of changing only boolean state variables, planning actions can decrease or increase a numeric variable. The goal state specifies constraints on this variable. In the simplest case the variable can only be increased by a static cost value in each action, and the goal state contains the objective to minimize the total cost. Systems such as metric-{\sc FF} \cite{hoffmann2003} do not guarantee optimality, but generally offer good results. We address our encoding of sentence generation with {\sc ptag} as metric planning in section \ref{ssec:pcrisp-domains}.

\subsection{Probabilistic TAG}
\label{ssec:probmodels}
{\sc ptag} \cite{resnik1992} views {\sc tag} derivations as sequences of probability events of three types, initial events, substitution events and adjunction events.

\newcommand{\adj}[0]{\textit{ adj}}
\newcommand{\init}[0]{\textit{ init}}
\newcommand{\subst}[0]{\textit{ subst}}
The probability distribution for initial events describes how likely it is to start any derivation with a given initial tree. 
It is defined over all initial trees $\alpha$ with their possible lexicalizations $w \in W_\alpha$: 
$$ \sum\limits_{\alpha \in I}\sum\limits_{w \in W_\alpha} P_{init}(\init(\alpha, w)) = 1 $$

For substitution events, there is a probability distribution for each substitution node $n$ of each initial tree $\tau$ lexicalized with $v$, which describes how likely it is to substitute it with an initial tree $\alpha$ lexicalized with $w$. 
$$\sum\limits_{\alpha \in I}\sum\limits_{w \in W_\alpha} P_{subst}(\subst(\tau,v , \alpha, w, n)) = 1$$

Similarly for each internal node there is a distribution that describes the probability to adjoin each auxiliary tree $\beta$ lexicalized with $w$. In addition some probability mass is reserved for the event of not adjoining anything to such a node at all.

$$ P_{adj}(\adj(\tau, v, \textit{None}, n)) +$$
$$ \sum\limits_{\beta \in A}\sum\limits_{w_\beta} P_{adj}(\adj(\tau, v, \beta, w, n)) = 1.$$

{\sc Ptag} assumes that all events occur independent of each other. Therefore it defines the total probability for a derivation as the product of the probability of its individual events.

\subsection{PCRISP Planning Domains}
\label{ssec:pcrisp-domains}
Using the definition of {\sc ptag}, we now reformulate the {\sc crisp} planning operators described in section \ref{ssec:crispdomain}.
The independence assumption in {\sc ptag} allows us to maintain that planning operator add a single tree, but are now assigned a probability score. However, while {\sc crisp} planning operators can add an elementary tree to any site of the right category, {\sc ptag} substitution and adjunction events are binary events between lexicalized trees at a specific node. We therefore adapt the literals that record open substitution and adjunction sites in partial derivations accordingly and create one operator for each node in each possible combination of lexicalized trees.
Figure \ref{example-action} shows an example planning operator for each type.
\begin{figure}[t]
\begin{center}
\cplanaction{\bf subst-t3-cat-t28-eats-n1(u,~x1)}{referent(u,~x1),\\ subst(t28-eats,~n1,~u), cat(x1)}{
$\lnot$needtoexpr(pred-cat,~x1),\\ $\lnot$subst(t-28-eats, n1,~u),\\
adj(t3-cat, n2 u)}{4.3012}\\\smallskip

@@

\caption{\label{example-action} {\sc pcrisp} Operator to substitute $t3$ lexicalized with 'cat' for node 2 of $t28$ lexicalized with `eats'.}
\end{center}
\end{figure}

Finally, we set the cost of an operator to be its negative log probability. For example 
\[
Cost( \subst(t_3,\textit{`cat'},t_{28},\textit{`eats"},n_2) ) =
\]
\[ 
 -\log( P(\subst(t_3,\textit{`cat'},t_{28},\textit{`eats"},n_2)) ). 
\] 
This way the plan with the smallest cost sum corresponds to the {\sc tag} derivation with the highest probability.

\subsection{Dealing with Data Sparseness}
\label{ssec:sparseness}
%To alleviate data sparseness, caused by the specific event definition of {\sc ptag}, we propose two alternative probability models. The first model is an unlexicalized version of {\sc ptag}, the second model uses a simple smoothing approach.\\ 
The probability event definition of {\sc ptag} is very fine grained. Substitution and adjunction events depend on specific parent and child trees with specific lexicalizations and on a node in the parent tree, as illustrated in figure \ref{fig:modelillustration}, B.1. When we estimate a probability model from training data, we cannot expect to observe evidence for all combinations of trees. Derivations that include such unseen events, have zero probability and are therefore impossible. As we show in section \ref{sec:experiments}, this gives rise to a massive data sparseness problem.

  A straightforward way to deal with data sparseness is to drop all lexicalizations from event definitions, as illustrated in figure \ref{fig:modelillustration}, A. Unfortunately this model does no longer account for bilexical dependencies between words. Since our system has to add a lexicalized tree in each step, lexicalizations for this child tree should always be taken into account by the probability model, if available. Despite these drawbacks, we perform experiments with the unlexicalized model as a baseline. This allows us to investigate if purely syntactic information is sufficient to achieve high quality generation output. \\  
An alternative model computes linear interpolation between three back-off levels. The first level is just standard {\sc ptag} (figure \ref{fig:modelillustration}, B.1), for the second level the lexicalizations of the parent tree is dropped (figure \ref{fig:modelillustration}, B.2), for the third level the model describes only the distribution of lexicalized child trees over each category (figure \ref{fig:modelillustration}, B.3).  
Notice also, that the third level is similar in spirit to a probabilistic version of original {\sc crisp} operators (compare figure \ref{fig:modelillustration}, B.3 and figure \ref{fig:crisp-operators}. 
\begin{figure}[t]
\begin{center}
\includegraphics[width=.5\textwidth]{figures/modelillustration}
\caption{\label{fig:modelillustration} Illustration of the unlexicalized probability model (A) and the three back-off levels of the linear interpolation model (B). B.1 is the original {\sc ptag} definition and corresponds directly to {\sc pcrisp} operators as in figure \ref{example-action}, while B.3 resembles {\sc crisp} operators (c.f. figure \ref{fig:crisp-operators}).}

@@

>>>>

We now extend \crisp\ to statistical generation (\pcrisp). The basic idea is to add a statistical grammar model while leaving the sentence generation mechanism untouched. This way we can select the highest scoring derivation which satisfies all constraints (grammaticality, expresses the communicative goal, uses unambiguous referring expressions, etc.).

@@

While we leave the basic sentence generation mechanism intact, we need to modify the concrete formulation of \crisp\ planning operators to accommodate bilexical dependencies. Likewise, we need to take the step from classical planning to \emph{metric} planning systems which can use the probabilities.
In metric planning \cite{fox2002}, planning actions can modify the value of numeric variables in addition to adding and deleting logical literals to the state. The goal state specifies constraints on this variable. In the simplest case the variable can only be increased by a static cost value in each action, and the goal state contains the objective to minimize the total cost. While systems such as Metric-{\sc FF} \cite{hoffmann2003} do not guarantee optimality, they do generally offer good results. We address our encoding of sentence generation with {\sc ptag} as metric planning in section \ref{ssec:pcrisp-domains}.

\subsection{Probabilistic TAG}
\label{ssec:probmodels}
{\sc ptag} \cite{resnik1992} views {\sc tag} derivations as sequences of events of three types: initial events, substitution events, and adjunction events.

The probability distribution for initial events describes how likely it is to start any derivation with a given initial tree. 
It is defined over all initial trees $\alpha \in I$ with their possible lexicalizations $w \in W_\alpha$: 
$$ \sum\limits_{\alpha \in I}\sum\limits_{w \in W_\alpha} P_i(\init(\alpha, w)) = 1 $$

For substitution events, there is a probability distribution for each substitution node $n$ of each elementary tree $\tau$ lexicalized with $v$, which describes how likely it is to substitute it with an initial tree $\alpha$ lexicalized with $w$. 
$$\sum\limits_{\alpha \in I}\sum\limits_{w \in W_\alpha} P_s(\subst(\tau,v , \alpha, w, n)) = 1$$

Similarly for each internal node there is a distribution that describes the probability to adjoin an auxiliary tree $\beta \in A$ lexicalized with $w$. In addition some probability mass is reserved for the event of not adjoining anything to such a node at all.

$$ P_a(\noadj(\tau, v, n)) + \strut$$
$$ \sum\limits_{\beta \in A}\sum\limits_{w \in W_\beta} P_a(\adj(\tau, v, \beta, w, n)) = 1.$$

{\sc Ptag} assumes that all events occur independently of each other. Therefore it defines the total probability for a derivation as the product of the probability of its individual events.

\subsection{PCRISP Planning Domains}
\label{ssec:pcrisp-domains}
Using the definition of {\sc ptag}, we now reformulate the \crisp\ planning operators described in section \ref{ssec:crispdomain}.
The independence assumption in {\sc ptag} allows us to continue to model each addition of a single elementary tree to the derivation (with a certain probability score). However, while \crisp\ planning operators can add an elementary tree to any site of the correct category, {\sc ptag} substitution and adjunction events are binary events between lexicalized trees at a specific node. We therefore adapt the literals that record open substitution and adjunction sites in partial derivations accordingly and create one operator for each node in each possible combination of lexicalized trees.
Fig.~\ref{example-action} shows an example planning operator for each type.
\begin{figure}[t]
\begin{center}
\cplanaction{\bf subst-t3-cat-t28-eats-n1(u,~x1)}{referent(u,~x1),\\ subst(t28-eats,~n1,~u), cat(x1)}{
$\lnot$needtoexpr(pred-cat,~x1),\\ $\lnot$subst(t-28-eats, n1,~u),\\
adj(t3-cat, n2 u)}{4.3012}\\\smallskip

@@

\caption{\label{example-action} Some \pcrisp\ operators for the grammar from Fig.~\ref{fig:grammar}.}
\end{center}
\end{figure}

Finally, we set the cost of an operator to be its negative log probability. For example 
\begin{align*}
&Cost(\mbox{subst-t3-cat-t28-eats-n1}) = \\
 & \qquad -\log P_s(\subst(\mathrm{t3},\textit{`cat'},\mathrm{t28},\textit{`eats'},\mathrm{n1})) ).
\end{align*}
This way the plan which minimizes the sum of the costs of its actions  corresponds to the {\sc tag} derivation with the highest probability.

\subsection{Dealing with Data Sparseness}
\label{ssec:sparseness}
The event definition of {\sc ptag} is very-fine grained. Substitution and adjunction events depend on specific parent and child trees with specific lexicalizations and on a node in the parent tree, as illustrated in Fig.~\ref{fig:modelillustration}, B.1. When we estimate a probability model from training data, we cannot expect to observe evidence for all combinations of trees. Derivations that include such unseen events have zero probability and are therefore impossible. As we show in section \ref{sec:experiments}, this gives rise to a massive data sparseness problem.

  A straightforward way to deal with data sparseness is to drop all lexicalizations from event definitions, as illustrated in Fig.~\ref{fig:modelillustration}, A. Unfortunately this model  no longer accounts for bilexical dependencies between words. Since our system has to add a lexicalized tree in each step, lexicalizations for this child tree should always be taken into account by the probability model, if available. Despite these drawbacks, we perform experiments with the unlexicalized model as a baseline. This allows us to investigate if purely syntactic information is sufficient to achieve high quality generation output.

An alternative model computes linear interpolation between three back-off levels. The first level is just standard {\sc ptag} (Fig.~\ref{fig:modelillustration}, B.1), for the second level the lexicalizations of the parent tree is dropped (Fig.~\ref{fig:modelillustration}, B.2), for the third level the model describes only the distribution of lexicalized child trees over each category (Fig.~\ref{fig:modelillustration}, B.3).  
Notice also that the third level is similar in spirit to a probabilistic version of original \crisp\ operators (compare Fig.~\ref{fig:modelillustration}, B.3 and Fig.~\ref{fig:crisp-operators}). 
\begin{figure}[t]
\begin{center}
\includegraphics[width=.5\textwidth]{figures/modelillustration}
\caption{\label{fig:modelillustration} Illustration of the unlexicalized probability model (A) and the three back-off levels of the linear interpolation model (B).}

@@

%%% Local Variables:  %%% mode: latex %%% TeX-master: "pcrisp-10" %%% End:

### 2010-05-21 trunk/Papers/pcrisp-10/related.tex
einmal ganz drueber

\section{Related Work}
\label{sec:related}
Statistical methods are extremely popular for surface realization, but have not been used in systems that integrate sentence planning. Most statistical generation approaches follow a generate-and-select strategy, first proposed by \newcite{knight1995} in their {\sc nitrogen} system. Such systems generate a set of candidate sentences using a conventional (non-statistical) grammar and then select the best output sentence by applying a statistical language model. This family includes systems such as {\sc HAL}ogen \cite{langkildeknight1998,langkilde2000} and open{\sc CCG} \cite{whitebaldridge2003}.  The {\sc Fergus} system \cite{bangalorerambow2000} is a variant of this approach that, like {\sc pcrisp}, employs {\sc ltag}. It first assigns elementary trees to each entity in the input sentence plan using a statistical tree model and then computes the most likely derivation using using only these trees with an $n$-gram model on the output sentence. An alternative to the generate and select approach is to use a probabilistic grammar model, like {\sc ptag}, trained on automatic parses \cite{zhongstent2005}. A related approach uses a model over local decisions of the generation system itself \cite{belz2008}. Both model can either be used to discriminate a set of output candidates, or more directly to choose the next best decision locally. Our approach is similar, because it uses {\sc ptag} to find the most likely output structure. However, the previous work discussed so far addresses surface realization only.
The generate and select approach is even incompatible with the integrated treatment of sentence planning and realization, because it discriminates between possible linearization of the same lexical material. In contrast, {\sc pcrisp} addresses sentence planning as well.

Our treatment of integrated sentence planning and surface realization as planning, is inherited directly from {\sc crisp} \cite{kollerstone2007}.  Planning has long played a role in generation, but has focused on discourse planning instead of specifically addressing sentence generation \cite{hovy1988,appelt1992}. The applicability of these ideas was limited at that time because efficient planning technology was not available. Recently the development of more efficient planning algorithms \cite{hoffmannnebel2001,richterwestphal2008} spawned a renewed interest in planning for NLG.  
{\sc Crisp} uses such algorithms to efficiently solve the sentence generation problem defined by {\sc spud} \cite{stonedoran1997}. {\sc Spud}, which instead used an incomplete greedy algorithm, is based on an {\sc ltag} whose trees are augmented with semantic and pragmatic constraints. Given a communicative goal, a solution to the {\sc spud} problem realizes this goal and simultaneously selects referring expressions. The next section explores {\sc crisp} in more detail.

>>>>

\section{Related Work}
\label{sec:related}

Statistical methods are popular for surface realization, but have not been used in systems that integrate sentence planning. Most statistical generation approaches follow a generate-and-select strategy, first proposed by \newcite{knight1995} in their {\sc nitrogen} system. Such systems generate a set of candidate sentences using a (possibly overgenerating) grammar and then select the best output sentence by applying a statistical language model. This family includes systems such as {\sc HAL}ogen \cite{langkildeknight1998,langkilde2000} and OpenCCG \cite{whitebaldridge2003}.  The {\sc Fergus} system \cite{bangalorerambow2000} is a variant of this approach which, like {\sc pcrisp}, employs {\sc ltag}. It first assigns elementary trees to each entity in the input sentence plan using a statistical tree model and then computes the most likely derivation using using only these trees with an $n$-gram model on the output sentence. An alternative to the generate and select approach is to use a probabilistic grammar model, like {\sc   ptag}, trained on automatic parses \cite{zhongstent2005}. A related approach uses a model over local decisions of the generation system itself \cite{belz2008}. Both models can either be used to discriminate a set of output candidates, or more directly to choose the next best decision locally. Our approach is similar in that it uses {\sc ptag} to find the most likely output structure. However, the previous work discussed so far addresses surface realization only. We extend this to a statistical NLG algorithm which does surface realization and sentence planning at the same time.

Our treatment of integrated sentence planning and surface realization as planning, is inherited directly from {\sc crisp} \cite{kollerstone2007}.  Planning has long played a role in generation, but has focused on discourse planning instead of specifically addressing sentence generation \cite{hovy1988,appelt1992}. The applicability of these ideas was limited at that time because efficient planning technology was not available. Recently the development of more efficient planning algorithms \cite{hoffmannnebel2001} spawned a renewed interest in planning for NLG.  {\sc Crisp} uses such algorithms to efficiently solve the sentence generation problem defined by {\sc spud} \cite{Stone2003a}. {\sc Spud}, which instead used an incomplete greedy algorithm, is based on an {\sc ltag} whose trees are augmented with semantic and pragmatic constraints. Given a communicative goal, a solution to the {\sc spud} problem realizes this goal and simultaneously selects referring expressions. The next section explores {\sc crisp} in more detail.

%%% Local Variables:  
%%% mode: latex 
%%% TeX-master: "pcrisp-10" 
%%% End:
